---
title: "BSAN 360 Final Project; Group 4"
author: "Group 4"
date: "9/2/2020"
output: html_document
---

# Titanic Survival Rate Data Analysis
#     Assima Imataliyeva, Tram Le, Michael Markin, Evie Nguyen, and Morgan Heller



##Original Data
```{r}

titanic <- read.csv(file.choose(), stringsAsFactors = FALSE)
View(titanic)

```

##Data Changes
```{r}

#Fill all the N/A Cabin values with “U” for unknown
titanic$Cabin[titanic$Cabin==" "] <- "U"
View(titanic)

#Determined some columns are irrelevent; removed them (PassengerID, Name, Ticket, Cabin)
titanic.std <- titanic[ ,c(-4,-9,-11)]
View(titanic.std)

#Fill in two missing values in "Embarked" column with "S" (the most common value)
titanic.std$Embarked[titanic.std$Embarked==" "] <- "S"
View(titanic.std)

#New vector for comparison (Family Size)
FamilySize <- (titanic.std$SibSp + titanic.std$Parch) + 1
titanic.std <- data.frame(titanic.std,FamilySize)
View(titanic.std)

# Define Family Size by categorical bins
for (i in 1:nrow(titanic.std)) {
  if (titanic.std$FamilySize[i] == 1) {
    titanic.std$FamilySize[i] <- 'Single'
  } else if (titanic.std$FamilySize[i] > 1 & titanic.std$FamilySize[i]<=3) {
    titanic.std$FamilySize[i] <- 'Small'
  } else if (titanic.std$FamilySize[i] > 3 & titanic.std$FamilySize[i]<=5) {
    titanic.std$FamilySize[i] <- 'Medium'
  } else if (titanic.std$FamilySize[i] > 5) {
    titanic.std$FamilySize[i] <- 'Large'}
}

#Remove empty Age rows
titanic.std <- with(titanic.std, titanic.std[!(Age == "" | is.na(Age)), ])

#Transform data types to factors and numeric
titanic.std$PassengerId <- as.numeric(titanic.std$PassengerId)
titanic.std$Age <- as.numeric(titanic.std$Age)
titanic.std$Fare <- as.numeric(titanic.std$Fare)
titanic.std$Pclass <- as.factor(titanic.std$Pclass)
titanic.std$Sex <- as.factor(titanic.std$Sex)
titanic.std$SibSp <- as.factor(titanic.std$SibSp)
titanic.std$Parch <- as.factor(titanic.std$Parch)
titanic.std$Embarked <- as.factor(titanic.std$Embarked)

#Scale Age and Fare columns
titanic.std[,c(5,8)] <- scale(titanic.std[,c(5,8)])

```

##Regression Models
#Model 1
```{r}

m1 <- lm(Survived ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked, data = titanic.std)
m1
summary(m1)

```

#Model 2; Optimized according to results from Model 1
```{r}

m3 <- lm(Survived ~ Pclass + Sex + Age +SibSp + Parch, data = titanic.std)
m3
summary(m3)

```

#Model 3; Multicollinear
```{r}

m2 <- lm(Survived ~ (Pclass + Sex + Age + SibSp + Parch + Fare + Embarked)^2, data = titanic.std)
summary(m2)

```

##Correlation
#Data Cleaning Specific to Correlation Calculations
```{r}

#Determined some columns are irrelevant; removed them (PassengerID, Name, Cabin, Ticket, Embarked)
titanic.corr <- titanic[,c(-1,-4,-9,-11)]


#Remove empty Age and Embarked rows
titanic.corr <- with(titanic.corr, titanic.corr[!(Age == "" | is.na(Age)), ])
titanic.corr <- with(titanic.corr, titanic.corr[!(Embarked == "" | is.na(Embarked)), ])

#Transform data types to numeric
titanic.corr$Survived <- as.numeric(titanic.corr$Survived)
titanic.corr$Age <- as.numeric(titanic.corr$Age)
titanic.corr$Fare <- as.numeric(titanic.corr$Fare)
titanic.corr$Pclass <- as.numeric(titanic.corr$Pclass)
titanic.corr$SibSp <- as.numeric(titanic.corr$SibSp)
titanic.corr$Parch <- as.numeric(titanic.corr$Parch)

#Re-label Sex and Embarked variables and transform to numeric
titanic.corr$Sex <- as.numeric(factor(titanic.corr$Sex, levels=c("male","female"), labels=c(0,1)))

titanic.corr$Embarked <- as.numeric(factor(titanic.corr$Embarked, levels=c("C","Q","S"), labels=c(1,2,3)))s

install.packages("corrplot")
library(corrplot)

#Create correlation matrix
cmat <- cor(titanic.corr)
corrplot.mixed(cmat)

```

##Visualization
#Box Plots
```{r}

library(ggplot2)

#Box plot fare against survived
ggplot(titanic, aes(x = as.factor(Survived), y = Fare)) +
  xlab("Survived")+
  ylab("Fare in Dollars")+
  ggtitle("Fare paid vs survival rate")+
  geom_boxplot()

#plot fare against class
ggplot(titanic, aes(x = as.factor(Pclass), y = Fare)) +
  xlab("Class")+
  ylab("Fare in Dollars")+
  ggtitle("Fare paid vs Class")+
  geom_boxplot()

#plot fare against embarked
ggplot(titanic, aes(x = as.factor(Embarked), y = Fare)) +
  xlab("Embarked")+
  ylab("Fare in Dollars")+
  ggtitle("Fare paid vs Embarked")+
  geom_boxplot()

```

#Bar Charts
```{r}

library(ggplot2)

#Relationship between sex and survival:
ggplot(data=titanic, aes(x=Sex,fill=as.factor(Survived)))+geom_bar()+
  xlab("Sex")+
  ylab("Number of Survived/Lost")+
  ggtitle("Bar Chart of Survival by Sex")+
  scale_fill_discrete(name="Legend",
                      breaks=c("0", "1"),
                      labels=c("Lost", "Survived"))

#...and by Pclass
ggplot(data=titanic, aes(x=Sex,fill=as.factor(Survived)))+geom_bar()+facet_grid(.~Pclass)+
  xlab("Sex")+
  ylab("Number of Survived/Lost")+
  ggtitle("Bar Chart of Survival by Sex by Passenger Class")+
  scale_fill_discrete(name="Legend",
                      breaks=c("0", "1"),
                      labels=c("Lost", "Survived"))


# Survival as a function of Pclass:
ggplot(data = titanic,aes(x=Pclass, fill = as.factor(Survived)))+
  geom_bar(position="fill")+
  ylab("Frequency")+
  xlab("Passenger Class")+
  ylab("Distribution of Survived/Lost by Passenger Class as a Percentage")+
  ggtitle("Bar Chart of Survival by Passenger Class")+
  scale_fill_discrete(name="Legend",
                      breaks=c("0", "1"),
                      labels=c("Lost", "Survived"))


# Relationship between age and survival by sex
ggplot(data = titanic, aes(x = Age, fill = as.factor(Survived))) + 
  geom_histogram() + 
  facet_grid(.~Sex)+
  xlab("Age")+
  ylab("Number of Survived/Lost")+
  ggtitle("Bar Chart of Survival by Sex by Age")+
  scale_fill_discrete(name="Legend",
                      breaks=c("0", "1"),
                      labels=c("Lost", "Survived"))


# Survival count by various Family Size categories
ggplot(data=titanic.std, aes(x=FamilySize, fill=as.factor(Survived))) +  
  geom_bar(stat="count") +
  xlab("Family Size") +
  ylab("Count of Survived/Dead") +
  ggtitle("Bar Chart of Survival by Family Sizes")

```

##Find Individual Metrics for Better Understanding
```{r}

#Look at fare, group by survived, find using function mean
tapply(titanic$Fare, titanic$Survived, FUN = mean)

#Look at fare, group by sex, find using function mean
tapply(titanic$Fare, titanic$Sex, FUN = mean)

#Look at fare, group by survived, find using function max
tapply(titanic$Fare, titanic$Survived, FUN = max)

#Look at fare, group by survived, find using function min
tapply(titanic$Fare, titanic$Survived, FUN = min)

#Look at fare, group by survived, pclass, and embarked, find using function mean
tapply(titanic$Fare, INDEX = titanic[c("Survived", "Pclass", "Embarked")], FUN=mean)

#Look at fare, group by survived and pclass, find using function mean
tapply(titanic$Fare, INDEX = titanic[c("Survived", "Pclass")], FUN=mean)

#Find the average survival age for each segment.
with(titanic,aggregate(formula=Age~Survived, FUN=mean))

#Using SibSp to group data for Survived
with(titanic,aggregate(formula=Survived~SibSp, FUN=mean))

#Look at survived, group by SibSp, find using function mean
tapply(titanic$Survived, titanic$SibSp, FUN = mean)

#Look at survived, group by passenger class, find using function mean
tapply(titanic$Survived, titanic$Pclass, FUN = mean)

```

